---
title: "PBL Draft"
author: "Andy, Blair, Julien"
output: pdf_document
date: "2026-02-27"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  quiet = TRUE,
  message = FALSE,
  warning = FALSE)
```

If needed, we kept all code + datasets + images at this link: https://github.com/jslee0605-miroh/CTA-Crime-ESDA

## Library Downloads

```{r}
library(sf)
library(RColorBrewer)
library(tmap)
library(viridis)
library(dplyr)
CRS.new <- st_crs("EPSG:3435")
```

## Data Retrieval

Reading in data for crime in 2025, CTA rail station locations, police station locations, and CTA ridership totals.
```{r}
crimes_raw <- st_read("crimes_2025.csv")
cta_rail_stations <- st_read("cta_rail_stations.csv")
police_station_raw <- st_read("police_stations.csv")
cta_ridership_raw <- st_read("cta_rail_ridership.csv")
```

We are using the Chicago community areas. 
```{r}
com_areas <- st_read("comm_area.geojson")
chicago_union <- st_union(com_areas) # using this to get rid of CTA stations outside of Chicago
```

## Data Cleaning/Wrangling

We need to filter CTA ridership data to just the year 2025. We also need to aggregate data, as ridership is currently broken down into months. 

```{r}
cta_ridership_clean <- cta_ridership_raw %>%
  mutate(YEAR = as.integer(YEAR)) %>% 
  filter(YEAR == 2025)

cta_ridership_clean <- cta_ridership_clean %>%
  mutate(TOTAL_RIDES = as.numeric(TOTAL_RIDES)) %>%  # ensure numeric
  group_by(RIDERSHIP_ID, NAME) %>%
  summarise(
    total_rides_2025 = sum(TOTAL_RIDES, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(desc(total_rides_2025))

head(cta_ridership_clean)
```

```{r}
head(crimes_raw)
```

We are also going to ignore crimes entries that have no location data (as otherwise, it is not possible to gauge whether it occurred within a buffer or not.) We also see that the datasets have latitude and longitude data but not point. We will convert to point data as well.

Crimes:
```{r}
crimes_clean <- crimes_raw %>% select(-Location)

crimes_clean <- crimes_clean %>%
  mutate(Latitude = as.numeric(Latitude), Longitude = as.numeric(Longitude))

crimes_clean <- crimes_clean %>% filter(!is.na(Latitude) & !is.na(Longitude)) # we go from 236549 observations to 236465

crimes_sf <- crimes_clean %>%
  st_as_sf(coords = c("Longitude", "Latitude"), crs = 4326, remove = FALSE)
```

Additionally, we will limit crime to those that directly impact people that would be utilizing the CTA stations. For example, we would consider battery but not vehicle theft.

```{r}
primary_type_values <- unique(crimes_sf$"Primary.Type")

primary_type_values
```

We will focus on crimes that impact public perception and create fear, such as robbery, battery, assault, etc.

```{r}
relevant_crimes <- c("SEX OFFENSE", "THEFT", "BATTERY", "MOTOR VEHICLE THEFT", "CRIMINAL SEXUAL ASSAULT", "ASSAULT", "WEAPONS VIOLATION", "ROBBERY", "PUBLIC PEACE VIOLATION", "NARCOTICS", "KIDNAPPING", "INTIMIDATION", "HOMICIDE", "PUBLIC INDECENCY", "HUMAN TRAFFICKING")

crimes_sf <- crimes_sf %>%
  filter(`Primary.Type` %in% relevant_crimes)
```


Ridership:
```{r}
cta_rail_stations_sf <- cta_rail_stations %>%
  st_as_sf(wkt = "the_geom", crs = 4326)

cta_rail_stations_sf <- cta_rail_stations_sf %>%
  st_filter(chicago_union, .predicate = st_within)
```

Police stations:
```{r}
police_station_clean <- police_station_raw %>% select(-LOCATION)

police_station_clean <- police_station_clean %>%
  mutate(LATITUDE = as.numeric(LATITUDE), LONGITUDE = as.numeric(LONGITUDE))

police_station_clean <- police_station_clean %>% filter(!is.na(LATITUDE) & !is.na(LONGITUDE)) # we go from 236549 observations to 236465

police_station_sf <- police_station_clean %>%
  st_as_sf(coords = c("LONGITUDE", "LATITUDE"), crs = 4326, remove = FALSE)

```

Transform to all use same CRS.
```{r}
com_areas <- st_transform(com_areas, CRS.new)
crimes_sf <- st_transform(crimes_sf, CRS.new)
cta_rail_stations_sf <- st_transform(cta_rail_stations_sf, CRS.new)
police_station_sf <- st_transform(police_station_sf, CRS.new)
```

```{r}
cta_rail_stations_sf <- cta_rail_stations_sf %>%
  left_join(cta_ridership_clean,
            by = c("STATION_ID" = "RIDERSHIP_ID"))
```

## ESDA

First, we plot police stations and CTA rail stations.

```{r}
tmap_mode("plot")

resources_map <- tm_shape(com_areas) + tm_borders() +
  tm_shape(police_station_sf) + tm_symbols(fill = "blue", shape = 24, size = 0.5) +
  tm_shape(cta_rail_stations_sf) + tm_symbols(fill = "red", size = 0.5) +
tm_layout(
    legend.outside = TRUE, 
    legend.outside.position = "right", 
    outer.margins = c(0.02, 0.02, 0.02, 0.18)) + 
tm_layout(
    main.title = paste0("Resource Points"),
    main.title.size = 1.5,
    main.title.position = c("center", "top"),
    legend.outside = TRUE,
    bg.color = "white",  
    frame = FALSE          
  ) +   
  tm_scalebar(position = c("left", "bottom"),
    text.size = 0.5 ) +
  tm_add_legend(
    type = "symbol",
    labels = "CTA Rail Stations",
    col = "red",
    shape = 16,
    size = 0.75
) +
  tm_add_legend(
    type = "symbol",
    labels = "Police Stations",
    col = "blue",
    shape = 17,
    size = 0.75
)
tmap_save(resources_map, "resources.png", width = 8, height = 8, units = "in", dpi = 300)
resources_map
```

Next, we want to go into ridership levels. 

```{r}
ridership_map <- tm_shape(com_areas) +
  tm_borders(col = "grey40") +
  tm_shape(cta_rail_stations_sf) +
  tm_symbols(
    size = "total_rides_2025",     
    col  = "total_rides_2025",   
    palette = "viridis",    
    style = "quantile",        
    alpha = 0.85,    
    border.col = "black",
    border.alpha = 0.2,
    title.size = "Ridership (2025)",
    title.col  = "Ridership (2025)"
  ) +
  tm_layout(
    main.title = "CTA Rail Station Ridership (2025)",
    legend.outside = TRUE,
    frame = FALSE
  ) +
  tm_scale_bar(position = c("left", "bottom"), size=2)

  tmap_save(ridership_map, "ridership_map.png", width = 9, height = 7, units = "in", dpi = 300)
  ridership_map
```


Buffers:
```{r}
buffer_quarter_mi <- st_buffer(cta_rail_stations_sf, 0.25 * 5280)
buffer_quarter_mi <- st_transform(buffer_quarter_mi, CRS.new)

buffer_half_mi <- st_buffer(cta_rail_stations_sf, 0.5 * 5280)
buffer_half_mi <- st_transform(buffer_half_mi, CRS.new)

buffer_1_mi <- st_buffer(cta_rail_stations_sf, 5280)
buffer_1_mi <- st_transform(buffer_1_mi, CRS.new)
```

Buffer plots:
```{r fig.height = 12}
tmap_mode("plot")
initial_buffer_map <- tm_shape(com_areas) + 
  tm_borders() + 
  tm_shape(cta_rail_stations_sf) + 
  tm_dots(fill="red", size=1) + 
  tm_shape(buffer_half_mi) + 
  tm_fill(col = "blue", alpha = 0.1) + 
  tm_borders(col = "blue") + 
  tm_layout(
    main.title = paste0("CTA Rail Stations: 0.25-mile Buffer"),
    main.title.size = 1.5,
    main.title.position = c("center", "top"),
    legend.outside = TRUE,
    bg.color = "white",    
    frame = FALSE     
  ) +   
  tm_scalebar(position = c("left", "bottom"),
    text.size = 1.5 )

  tmap_save(initial_buffer_map, "initial_buffer_map.png", width = 8, height = 6, units = "in", dpi = 300)
  initial_buffer_map
```

Crime count in buffer:
```{r}
crime_025_buffer <- lengths(st_intersects(buffer_quarter_mi, crimes_sf))
crime_05_buffer <- lengths(st_intersects(buffer_half_mi, crimes_sf))
crime_1_buffer <- lengths(st_intersects(buffer_1_mi, crimes_sf))

cta_rail_stations_sf$crime_count_025 <- crime_025_buffer
cta_rail_stations_sf$crime_count_05  <- crime_05_buffer
cta_rail_stations_sf$crime_count_1  <- crime_1_buffer

#IMPORTANT: overlapping buffers count crime twice
```

Police station count in buffer:

```{r}
police_025_buffer <- lengths(st_intersects(buffer_quarter_mi, police_station_sf))
police_05_buffer <- lengths(st_intersects(buffer_half_mi, police_station_sf))
police_1_buffer <- lengths(st_intersects(buffer_1_mi, police_station_sf))

cta_rail_stations_sf$police_count_025 <- police_025_buffer
cta_rail_stations_sf$police_count_05 <- police_05_buffer
cta_rail_stations_sf$police_count_1 <- police_1_buffer

# From this, only 1 CTA stations has more than 1 police station within a mile radius
# will use 1 mile buffer for police station analysis
```


Crime Exposure Visualization:
```{r}
cta_rail_stations_sf <- cta_rail_stations_sf %>%
  mutate(
    crime_level_025 = case_when(
      crime_count_025 >= quantile(crime_count_025, 0.75, na.rm = TRUE) ~ "High",
      crime_count_025 >= quantile(crime_count_025, 0.25, na.rm = TRUE) ~ "Medium",
      TRUE ~ "Low"
    ),
    crime_level_025 = factor(crime_level_025, levels = c("Low", "Medium", "High"))
  )

exposure_025_buffer <- st_buffer(cta_rail_stations_sf, 0.25 * 5280)
```

```{r}
cta_rail_stations_sf <- cta_rail_stations_sf %>%
  mutate(
    crime_level_05 = case_when(
      crime_count_05 >= quantile(crime_count_05, 0.8, na.rm = TRUE) ~ "Very High",
      crime_count_05 >= quantile(crime_count_05, 0.6, na.rm = TRUE) ~ "High",
      crime_count_05 >= quantile(crime_count_05, 0.4, na.rm = TRUE) ~ "Moderate",
      crime_count_05 >= quantile(crime_count_05, 0.2, na.rm = TRUE) ~ "Low",
      crime_count_05 >= quantile(crime_count_05, 0, na.rm = TRUE) ~ "Very Low",
    ),
    crime_level_05 = factor(crime_level_05, levels = c("Very High", "High", "Moderate", "Low", "Very Low"))
  )

exposure_05_buffer <- st_buffer(cta_rail_stations_sf, 0.5 * 5280)
```

First, we create a map to see how many relevant crimes are committed in the 0.5 mile buffer.

```{r}
tmap_mode("plot")
crime_count_map <- tm_shape(com_areas) + 
  tm_borders() + 
  tm_shape(cta_rail_stations_sf) + 
  tm_dots(fill="black", size=0.25) + 
  tm_shape(exposure_05_buffer) + 
  tm_fill(
    col = "crime_level_05",
    palette = rev(brewer.pal(5, "YlOrRd")),
    title = "Crime Count",
      border.col = "black",
    border.alpha = 0.2,
    alpha = 0.5
  ) + 
  tm_layout(
    main.title = paste0("CTA Rail Stations Crime Count: 0.5-mile Buffer"),
    main.title.size = 1.5,
    main.title.position = c("center", "top"),
    legend.outside = TRUE,
    bg.color = "white",    
    frame = FALSE     
  ) +   
  tm_scalebar(position = c("left", "bottom"),
    text.size = 0.5 )

tmap_save(crime_count_map, "crime_count_map.png", width = 8, height = 6, units = "in", dpi = 300)
crime_count_map
```

```{r}
cta_rail_stations_sf <- cta_rail_stations_sf %>%
  mutate(
    # rate per 10,000 riders (avoid divide-by-zero)
    crime_rate_05 = ifelse(total_rides_2025 > 0,
                           (crime_count_05 / total_rides_2025) * 10000,
                           NA_real_),

    # 5-quantile bins (20% each)
    crime_rate_level_05 = cut(
      crime_rate_05,
      breaks = quantile(crime_rate_05, probs = seq(0, 1, 0.2), na.rm = TRUE),
      include.lowest = TRUE,
      labels = c("Very Low", "Low", "Moderate", "High", "Very High"),
    ),
    
    crime_rate_level_05 = factor(
      crime_rate_level_05,
      levels = c("Very High", "High", "Moderate", "Low", "Very Low"),
      ordered = TRUE
    )
  )

exposure_05_buffer <- st_buffer(cta_rail_stations_sf, 0.5 * 5280)
```

In an attempt to see how many relevant crimes were being committed relative to the number of people actually using the stop, we also created a map for crime rate, measured by raw crime count divided by ridership at that given station.

```{r}
tmap_mode("plot")
crime_rate_map <- tm_shape(com_areas) + 
  tm_borders() + 
  tm_shape(cta_rail_stations_sf) + 
  tm_dots(fill="black", size=0.25) + 
  tm_shape(exposure_05_buffer) + 
  tm_fill(
    col = "crime_rate_level_05",
    palette = rev(brewer.pal(5, "YlOrRd")),
    title = "Crime Rate",
    border.col = "black",
    border.alpha = 0.2,
    alpha = 0.5
  ) + 
  tm_layout(
    main.title = paste0("CTA Rail Stations Crime Rate: 0.5-mile Buffer"),
    main.title.size = 1.5,
    main.title.position = c("center", "top"),
    legend.outside = TRUE,
    bg.color = "white",    
    frame = FALSE     
  ) +   
  tm_scalebar(position = c("left", "bottom"),
    text.size = 0.5 )

tmap_save(crime_rate_map, "crime_rate_map.png", width = 8, height = 6, units = "in", dpi = 300)
crime_rate_map
```
Filtering for Police Stations within a 1 mile radius
```{r}
cta_with_police <- cta_rail_stations_sf %>%
    filter(police_count_1 >= 1)

exposure_05_buffer_police <- st_buffer(cta_with_police, 0.5 * 5280)

cta_wo_police <- cta_rail_stations_sf %>%
    filter(police_count_1 == 0)

exposure_05_buffer_no_police <- st_buffer(cta_wo_police, 0.5 * 5280)
```

Crime exposure map filtered for police stations nearby:
```{r}
tmap_mode("plot")
police_station_nearby_map <- tm_shape(com_areas) + 
  tm_borders() + 
  tm_shape(cta_rail_stations_sf) + 
  tm_dots(fill="black", size=0.25) + 
  tm_shape(exposure_05_buffer_police) + 
  tm_fill(
    col = "crime_level_05",
    palette = rev(brewer.pal(5, "YlOrRd")),
    title = "Crime Level",
        border.col = "black",
    border.alpha = 0.2,
    alpha = 0.5
  ) + 
  tm_layout(
    main.title = paste0("CTA Rail Stations Crime Exposure: 0.5-mile Buffer with a Police Station Nearby"),
    main.title.size = 1.5,
    main.title.position = c("center", "top"),
    legend.outside = TRUE,
    bg.color = "white",    
    frame = FALSE     
  ) +   
  tm_scalebar(position = c("left", "bottom"),
    text.size = 0.5 )

tmap_save(police_station_nearby_map, "police_station_nearby_map.png", width = 8, height = 6, units = "in", dpi = 300)
police_station_nearby_map
```
Crime Exposure Map w/o police stations nearby:
```{r}
tmap_mode("plot")
police_station_not_nearby_map <- tm_shape(com_areas) + 
  tm_borders() + 
  tm_shape(cta_rail_stations_sf) + 
  tm_dots(fill="black", size=0.25) + 
  tm_shape(exposure_05_buffer_no_police) + 
  tm_fill(
    col = "crime_level_05",
    palette = rev(brewer.pal(5, "YlOrRd")),
    title = "Crime Level",
        border.col = "black",
    border.alpha = 0.2,
    alpha = 0.5
  ) + 
  tm_layout(
    main.title = paste0("CTA Rail Stations Crime Exposure: 0.5-mile Buffer without a Police Station Nearby"),
    main.title.size = 1.5,
    main.title.position = c("center", "top"),
    legend.outside = TRUE,
    bg.color = "white",    
    frame = FALSE     
  ) +   
  tm_scalebar(position = c("left", "bottom"),
    text.size = 0.5 )

tmap_save(police_station_not_nearby_map, "police_station_not_nearby_map.png", width = 8, height = 6, units = "in", dpi = 300)
police_station_not_nearby_map
```


```{r}
nearest_police_idx <- st_nearest_feature(cta_rail_stations_sf, police_station_sf)

# compute minimum distance
min_dist <- st_distance(
  cta_rail_stations_sf,
  police_station_sf[nearest_police_idx, ],
  by_element = TRUE
)

cta_rail_stations_sf <- cta_rail_stations_sf %>%
  mutate(
    min_dist_police = as.numeric(min_dist),
    min_dist_police_miles = min_dist_police / 5280
  )

```

```{r}
tmap_mode("plot")

dist_to_police <- tm_shape(com_areas) +
  tm_borders() +
tm_shape(cta_rail_stations_sf) +
  tm_dots(
    fill = "min_dist_police_miles",          # <-- map fill, not col
    fill.scale = tm_scale_intervals(
      n = 5,
      style = "quantile",
      values = "-Oranges"
    ),
    fill.legend = tm_legend(title = "Min Distance (mi)"),
    size = 0.5
  ) +
tm_layout(
  main.title = "CTA Stations: Distance to Nearest Police Station",
  legend.outside = TRUE,
  frame = FALSE
) +
tm_scalebar(position = c("left", "bottom"), text.size = 0.5)

tmap_save(dist_to_police, "dist_to_police.png", width = 8, height = 6, units = "in", dpi = 300)
dist_to_police
```


## Non-Spatial EDA

Regression with Crime Count 05 and 2025 Ridership

```{r}
png("ridership_vs_crimecount.png", width = 2000, height = 1500, res = 300)
ridership_v_crimecount <- plot(cta_rail_stations_sf$crime_count_05,
     cta_rail_stations_sf$total_rides_2025,
     pch = 19,
     col = "darkgray",
     xlab = "Crime Count within 0.5 Miles",
     ylab = "Ridership (2025)",
     main = "CTA 2025 Ridership vs Crime Count")
model <- lm(total_rides_2025 ~ crime_count_05, data = cta_rail_stations_sf)
abline(model, col = "blue", lwd = 2)
model_sum = summary(model)
r2 = model_sum$r.squared
slope <- model_sum$coefficients[2, 1]
p_value <- model_sum$coefficients[2, 4]
legend("topright",
       legend = c(
         paste0("R^2 = ", round(r2, 3)),
         paste0("Slope = ", round(slope, 3)),
         paste0("p = ", signif(p_value, 3))
       ),
       bty = "n")
dev.off()
ridership_v_crimecount
```
Scatter Plot for Minimum distance vs Crime Count
```{r}
png("mindist_v_crimerate.png", width = 2000, height = 1500, res = 300)
plot(cta_rail_stations_sf$min_dist_police_miles,
     cta_rail_stations_sf$crime_rate_05,
     pch = 19,
     col = "darkgray",
     xlab = "CTA Station Distance from Nearest Police Station",
     ylab = "Crime Rate within 0.5 miles",
     main = "Crime Rate vs. Police Station Minimum Distance")
axis(2, at = 0:3, labels = 0:3)
model <- lm(crime_rate_05 ~ min_dist_police_miles, data = cta_rail_stations_sf)
abline(model, col = "blue", lwd = 2)
model_sum = summary(model)
r2 = model_sum$r.squared
slope <- model_sum$coefficients[2, 1]
p_value <- model_sum$coefficients[2, 4]
legend("topright",
       legend = c(
         paste0("R^2 = ", round(r2, 3)),
         paste0("Slope = ", round(slope, 3)),
         paste0("p = ", signif(p_value, 3))
       ),
       bty = "n")
dev.off()
```


